% This file is part of Oaklisp.
%
% This program is free software; you can redistribute it and/or modify
% it under the terms of the GNU General Public License as published by
% the Free Software Foundation; either version 2 of the License, or
% (at your option) any later version.
%
% This program is distributed in the hope that it will be useful,
% but WITHOUT ANY WARRANTY; without even the implied warranty of
% MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
% GNU General Public License for more details.
%
% The GNU GPL is available at http://www.gnu.org/licenses/gpl.html
% or from the Free Software Foundation, 59 Temple Place - Suite 330,
% Boston, MA 02111-1307, USA


\chapter{Administrative Details}

\section{Getting a Copy}

The most recent released version of Oaklisp, along with the manuals,
are available at \url{http://www.bcl.hamilton.ie/~barak/oaklisp/}, but
this is to be superseded by a new site:
\url{http://oaklisp.alioth.debian.org/}.

\section{Bugs}

The following are known serious problems and inadequacies of the
current implementation.  People are invited to work on remedying them.
None of these are fundamental; they're simply due to lack of either
effort or motivation.

\begin{itemize}
\item Floating point numbers are not supported.  Rationals can be
used to make up for this lack.

\item In contrast to the error handling system, which is Industrial
Strength, the debugger barely exists.

\item There is no foreign function interface for loading and calling
C routines from a running Oaklisp.

%% Removed because this has been tightened up so much that only someone
%% familiar with the internals would be able to cobble up something that
%% would cause a core dump when invoked, and that would take some work.
%
% \item Calling some non-operations dumps core rather than invoking the
% debugger.
\end{itemize}

Bug reports, enhancements, and the like should be posted using the
facilities on \url{http://oaklisp.alioth.debian.org/}; queries can
also be sent to \texttt{barak+oaklisp@pearlmutter.net}.

We appreciate enhancements (especially in the form of patch files),
bug fixes, and bug reports.  We are particularly grateful for porting
problem fixes.  In a bug report, please include the precise version of
Oaklisp, which is indicated by the date at the end of the tar file.
And please try to make sure that it's really a bug and not a feature,
and pretty please, if at all possible, find a \emph{very short} program
that manifests your bug.  In any case please be aware that we are
under no obligation to respond to bug reports in any way whatsoever.

\section{Copyright and Lack of Warranty}

The Oaklisp copyright belongs to its authors.  It is authorized for
distribution under the GNU General Public License, version 2, copies
of which are readily obtainable from the Free Software Foundation.
There is no warranty; use at your own risk.  For more precise
information, see the COPYING file in the Oaklisp source distribution.
